\vspace{-0.5in}{\bfseries\scshape Transfer Learning for computationally limited training of Vision Transformers on art classification tasks } \\ \vspace{1ex} \normalsize{Bachelor's Project Thesis}\vspace{-2ex}

### 2022-06-30 writing/tex/draft_introduction.tex
First part of the presentation pretty much done. Now second part: my methods and results
Continued working on writing y paper

\section{Introduction}
Since the introduction of AlexNet, roughly a decade ago, \textit{Convolutional Neural Networks} (CNNs) have played an important role in \textit{Computer Vision} (CV) \citep{krizhevsky2012imagened}. Recently, however, a new type of architecture, called \textit{Vision Transformer} (ViT), gained state of the art performance on common learning benchmarks, including the ImageNet dataset\footnote{At the time of writing TODO is the best performing model, and the top TODO are all Transformer-based: (website)}.

Exciting as this may be, a plain ViT is not likely to become useful in circumstances with limited training data and/or only consumer-grade hardware available for learning. This is due to a limitation inherent to the ViT architecture, which will be elaborated on in section \ref{vits}. Consequently, the current paper investigates whether utilizing \textit{Transfer Learning} (TL) capabilities of ViTs can help overcome these limitations, and if this allows ViTs to become the preferred architecture in the aforementioned circumstances as well.

Before going into more detail, however, some background information is needed. The remainder of this section will first briefly describe CNNs and ViTs, and compare them to one another. What follows, is an explanation of TL, and why it might help ViTs in circumstances with limited data/compute available. Finally, related work is discussed, and a research question is proposed.

%\begin{figure}[!tbp]
%    \centering
%    \includegraphics[width=7.7cm]{img/vit_vs_cnn.png}
%    \caption{Intuition behind CNNs and ViTs. Note that many details are left out for the purpose of illustration. The top half compares MLP layers (left) to those of CNNs (right). A single neuron and its receptive field are highlighted. Connections are only allowed such as the accentuated 4, which reduces the total number of connections greatly. Moreover, the non-accentuated connections don't add to the number of trainable weights, as the same 4 are shared between all neuron-receptive-field-pairs.\\The bottom half draws a similar comparison between MLPs (left) and ViTs (right). A single query vector is combined with all key vectors to get a scalar value for each position (blue boxes). These are multiplied with corresponding value vectors (in red), and then added to get a weighted sum (green). This sum is fed through one or more MLP-type layers, to produce a single output position (opaque). The same steps are taken for all other positions (transparent) as well, using their corresponding query vectors.}
%    \label{vitVsCnn}
%\end{figure}

@@

\subsection{Vision Transformers} \label{vits}
\paragraph{Architecture}
ViTs were first introduced in \citeauthor{dosovitskiy2020image} (\citeyear{dosovitskiy2020image}), and are based on the \textit{encoder stack} of the \textit{Transformer} architecture \citep{vaswani2017attention}. This stack takes a sequence of vectors as input, and is composed of alternating \textit{multi-headed self-attention} layers and position-wise MLP-type layers -- position-wise meaning the MLP can take only a single position (constituent vector) as input at a time, and is shared between all positions.

The self-attention mechanism plays a central role in ViTs, and is worth elaborating on. Each head of a self-attention layer contains 3 learned projection matrices, which are used to produce $key$, $value$, and $query$ vectors for every position in the input sequence. The output for a single position is a weighted sum of all $value$ vectors. Weights are obtained by matching this position's $query$ with all $key$ vectors\footnote{In principle by taking softmax over all query-key inner products.}. As such, an attention head decides itself how much of each position should be incorporated in subsequent representations.

The original Transformer architecture was designed to solve problems in \textit{natural language processing} (a field it now dominates). To make it work for CV applications, ViTs divide an image into square patches, which get treated as tokens (words). Learned embeddings convert these into a sequence of vectors, to which 1D positional encodings are added\footnote{Do mind that ViTs have no built-in knowledge of an image's 2D topology.}. A learned $[class]$ vector gets prepended to the sequence, and its representation at the end of the encoder stack is the final output, which can then be used for classification, segmentation, etc.

@@

%ViTs were first introduced in \citeauthor{dosovitskiy2020image} (\citeyear{dosovitskiy2020image}), and are based on the \textit{Transformer} architecture \citep{vaswani2017attention}. This type of ANN was originally designed to solve problems in \textit{natural language processing} (a field it now dominates). It takes a series of tokens as input, and uses a learned mapping to convert them into a series of vectors, called \textit{token embeddings}. These embeddings are then fed through an encoder and decoder, but since the latter is discarded for ViTs, it won't be discussed here.

%, and takes a variable length sequence of \textit{token embeddings}\footnote{The model has some vocabulary of tokens it knows, for example all words in the English language. For each it learns a vector representation, called a token embedding.} as input. While Transformers are made up of an \textit{encoder} and \textit{decoder}, the latter won't be discussed here, as it is discarded in the ViT architecture.

@@

%The architecture consists of an \textit{encoder}, and \textit{decoder}, but since the latter is discarded in ViTs, it won't be discussed here.

@@

% While a CNN is restricted to only incorporate information from its surrounding pixels in a subsequent layer, a ViT

\paragraph{CNNs versus ViTs}
Where CNNs reduce the number of trainable weights by taking an image's topology into account, ViTs do this by learning themselves what information they should incorporate in successive representations of the input. As such, they are not restricted to a local receptive field, but can potentially include information from all over the previous layer.

Moreover, weight sharing applies as much to ViTs as it does to CNNs. In ViTs, each position goes through the same computational pipelines of being multiplied by 3 projection matrices, followed by a weighted sum calculation, followed by being fed through an MLP-type layer\footnote{Neglecting skip connections and normalization, as this is a comparison on an intuitive level.}. In CNNs these pipelines take the form of convolution kernels.

\paragraph{A limitation of ViTs}
While ViTs are promising; the fact that they are not tailored to CV tasks, makes that they lack a so called image specific \textit{Inductive Bias}, which CNNs do have. Intuitively this means a ViT first has to learn what an image is, before it can move on to the learning task at hand. The result is that they require substantially more training data (and time) than CNNs do.

It is for this reason the paper that first introduced ViTs \citep{dosovitskiy2020image} mentions using pre-trained models rather than randomly initialised ones, as it allows general knowledge about images to be carried over to the new task. This technique falls within the category of TL, which will be covered next.

% CNNs have image specific inductive bias: the architecture itself is tailored to 2D images, and as such has structural information about how images work built-in. This is not the case for ViTs, which lack an image specific inductive bias, which makes them very data hungry.

@@

\subsection{Related work and the current study} \label{related_work}
As mentioned, the paper that first introduced ViTs already demonstrated they benefit from TL \citep{dosovitskiy2020image}, and more has been discovered about ViTs' TL properties since then. \citeauthor{matsoukas2021time} (\citeyear{matsoukas2021time}) investigated whether TL allowed ViTs to replace CNNs in Medical Imaging. This domain is characterized by small datasets, so to overcome the inductive bias problem described earlier, TL is imperative. It was found that ViTs benefit more from FT TL than CNNs do, and that ViTs pre-trained on ImageNet perform on par CNNs.

What \citeauthor{matsoukas2021time} did not investigate, is OTS performance of ViTs. This was, however, done in \citeauthor{zhou2021convnets} (\citeyear{zhou2021convnets}), which found that ViTs pre-trained on ImageNet had worse OTS performance than their CNN counterparts. An exception was the \textit{Swin-B} architecture, which was the best performing model on all datasets, and uses \textit{(shifted-)window-based} self-attention methods, to make Transformers better suited for CV tasks \citep{liu2021swin}.
% Not into detail about ots and other techniques

Another ``shortcomming'' of \citeauthor{matsoukas2021time} is that it merely examined whether ViTs can use TL to outperform CNNs in a domain that happens to be characterized by limited-sized datasets, but that it did not actively focus on the extent to which ViTs can be made useful in limited data/compute environments. More related to this is \citeauthor{touvron2021training} (\citeyear{touvron2021training}), which proposed transformer-specific distillation-based learning techniques to make ViTs applicable for smaller datasets. This is not necessarily related to TL as defined in this paper, however.
% matsoukas2021time not specifically focussed on seeing to what degree TL allows ViTs to dominate CNNs in limited compute/data environments
\\\\
The current study will examine to what extent TL techniques (both FT and OTS) allow ViTs to become the preferred architecture when compute resources are limited, and attempts to find a methodology that is recommended in these circumstances. It will, however, restrict itself to art classification tasks. Several experiments will be performed, of which results will be analysed both quantitatively as well as qualitatively.

%The current study limits itself to the task of identifying which materials were used in artworks. For this, it will examine to what extent TL techniques (both fine-tuning and OTS) allow ViTs to become the preferred architecture when compute resources are limited, and attempt to find a methodology that is preferred in circumstances as described. It will do this by means of several experiments, and both quantitative and qualitative analysis of the results.

More formally, the following research question is proposed: \begin{quote}\textit{To what extent (and in what manners) does utilizing TL allow ViTs to perform on par with CNNs on common art classification problems, when compute resources are limited to a single GPU?}\end{quote}

>>>>

\section{Introduction}
Since the introduction of AlexNet, roughly a decade ago, \textit{Convolutional Neural Networks} (CNNs) have played an important role in \textit{Computer Vision} (CV) \citep{krizhevsky2012imagened}. Recently, however, a new type of architecture, called \textit{Vision Transformer} (VT), gained state of the art performance on common learning benchmarks, including the ImageNet dataset\footnote{At the time of writing TODO is the best performing model, and the top TODO are all Transformer-based: (website)}.

Exciting as this may be, a plain VT is not likely to become useful in circumstances with limited training data and/or only consumer-grade hardware available for learning. This is due to a limitation inherent to the VT architecture, which will be elaborated on in section \ref{vits}. Consequently, the current paper investigates whether utilizing \textit{Transfer Learning} (TL) capabilities of VTs can help overcome these limitations, and if this allows VTs to become the preferred architecture in the aforementioned circumstances as well.

Before going into more detail, however, some background information is needed. The remainder of this section will first briefly describe CNNs and VTs, and compare them to one another. What follows, is an explanation of TL, and why it might help VTs in circumstances with limited data/compute available. Finally, related work is discussed, and a research question is proposed.

%\begin{figure}[!tbp]
%    \centering
%    \includegraphics[width=7.7cm]{img/vit_vs_cnn.png}
%    \caption{Intuition behind CNNs and VTs. Note that many details are left out for the purpose of illustration. The top half compares MLP layers (left) to those of CNNs (right). A single neuron and its receptive field are highlighted. Connections are only allowed such as the accentuated 4, which reduces the total number of connections greatly. Moreover, the non-accentuated connections don't add to the number of trainable weights, as the same 4 are shared between all neuron-receptive-field-pairs.\\The bottom half draws a similar comparison between MLPs (left) and VTs (right). A single query vector is combined with all key vectors to get a scalar value for each position (blue boxes). These are multiplied with corresponding value vectors (in red), and then added to get a weighted sum (green). This sum is fed through one or more MLP-type layers, to produce a single output position (opaque). The same steps are taken for all other positions (transparent) as well, using their corresponding query vectors.}
%    \label{vitVsCnn}
%\end{figure}

@@

\subsection{Vision Transformers} \label{vits}
\paragraph{Architecture}
VTs were first introduced in \citeauthor{dosovitskiy2020image} (\citeyear{dosovitskiy2020image}), and are based on the \textit{encoder stack} of the \textit{Transformer} architecture \citep{vaswani2017attention}. This stack takes a sequence of vectors as input, and is composed of alternating \textit{multi-headed self-attention} layers and position-wise MLP-type layers -- position-wise meaning the MLP can take only a single position (constituent vector) as input at a time, and is shared between all positions.

The self-attention mechanism plays a central role in VTs, and is worth elaborating on. Each head of a self-attention layer contains 3 learned projection matrices, which are used to produce $key$, $value$, and $query$ vectors for every position in the input sequence. The output for a single position is a weighted sum of all $value$ vectors. Weights are obtained by matching this position's $query$ with all $key$ vectors\footnote{In principle by taking softmax over all query-key inner products.}. As such, an attention head decides itself how much of each position should be incorporated in subsequent representations.

The original Transformer architecture was designed to solve problems in \textit{natural language processing} (a field it now dominates). To make it work for CV applications, VTs divide an image into square patches, which get treated as tokens (words). Learned embeddings convert these into a sequence of vectors, to which 1D positional encodings are added\footnote{Do mind that VTs have no built-in knowledge of an image's 2D topology.}. A learned $[class]$ vector gets prepended to the sequence, and its representation at the end of the encoder stack is the final output, which can then be used for classification, segmentation, etc.

@@

%VTs were first introduced in \citeauthor{dosovitskiy2020image} (\citeyear{dosovitskiy2020image}), and are based on the \textit{Transformer} architecture \citep{vaswani2017attention}. This type of ANN was originally designed to solve problems in \textit{natural language processing} (a field it now dominates). It takes a series of tokens as input, and uses a learned mapping to convert them into a series of vectors, called \textit{token embeddings}. These embeddings are then fed through an encoder and decoder, but since the latter is discarded for VTs, it won't be discussed here.

%, and takes a variable length sequence of \textit{token embeddings}\footnote{The model has some vocabulary of tokens it knows, for example all words in the English language. For each it learns a vector representation, called a token embedding.} as input. While Transformers are made up of an \textit{encoder} and \textit{decoder}, the latter won't be discussed here, as it is discarded in the VT architecture.

@@

%The architecture consists of an \textit{encoder}, and \textit{decoder}, but since the latter is discarded in VTs, it won't be discussed here.

@@

% While a CNN is restricted to only incorporate information from its surrounding pixels in a subsequent layer, a VT

\paragraph{CNNs versus VTs}
Where CNNs reduce the number of trainable weights by taking an image's topology into account, VTs do this by learning themselves what information they should incorporate in successive representations of the input. As such, they are not restricted to a local receptive field, but can potentially include information from all over the previous layer.

Moreover, weight sharing applies as much to VTs as it does to CNNs. In VTs, each position goes through the same computational pipelines of being multiplied by 3 projection matrices, followed by a weighted sum calculation, followed by being fed through an MLP-type layer\footnote{Neglecting skip connections and normalization, as this is a comparison on an intuitive level.}. In CNNs these pipelines take the form of convolution kernels.

\paragraph{A limitation of VTs}
While VTs are promising; the fact that they are not tailored to CV tasks, makes that they lack a so called image specific \textit{Inductive Bias}, which CNNs do have. Intuitively this means a VT first has to learn what an image is, before it can move on to the learning task at hand. The result is that they require substantially more training data (and time) than CNNs do.

It is for this reason the paper that first introduced VTs \citep{dosovitskiy2020image} mentions using pre-trained models rather than randomly initialised ones, as it allows general knowledge about images to be carried over to the new task. This technique falls within the category of TL, which will be covered next.

% CNNs have image specific inductive bias: the architecture itself is tailored to 2D images, and as such has structural information about how images work built-in. This is not the case for VTs, which lack an image specific inductive bias, which makes them very data hungry.

@@

\subsection{Related work and the current study} \label{related_work}
As mentioned, the paper that first introduced VTs already demonstrated they benefit from TL \citep{dosovitskiy2020image}, and more has been discovered about VTs' TL properties since then. \citeauthor{matsoukas2021time} (\citeyear{matsoukas2021time}) investigated whether TL allowed VTs to replace CNNs in Medical Imaging. This domain is characterized by small datasets, so to overcome the inductive bias problem described earlier, TL is imperative. It was found that VTs benefit more from FT TL than CNNs do, and that VTs pre-trained on ImageNet perform on par CNNs.

What \citeauthor{matsoukas2021time} did not investigate, is OTS performance of VTs. This was, however, done in \citeauthor{zhou2021convnets} (\citeyear{zhou2021convnets}), which found that VTs pre-trained on ImageNet had worse OTS performance than their CNN counterparts. An exception was the \texttt{Swin} architecture, which was the best performing model on all datasets, and uses \textit{(shifted-)window-based} self-attention methods, to make Transformers better suited for CV tasks \citep{liu2021swin}.
% Not into detail about ots and other techniques

Another ``shortcomming'' of \citeauthor{matsoukas2021time} is that it merely examined whether VTs can use TL to outperform CNNs in a domain that happens to be characterized by limited-sized datasets, but that it did not actively focus on the extent to which VTs can be made useful in limited data/compute environments. More related to this is \citeauthor{touvron2021training} (\citeyear{touvron2021training}), which proposed transformer-specific distillation-based learning techniques to make VTs applicable for smaller datasets. This is not necessarily related to TL as defined in this paper, however.
% matsoukas2021time not specifically focussed on seeing to what degree TL allows VTs to dominate CNNs in limited compute/data environments
\\\\
The current study will examine to what extent TL techniques (both FT and OTS) allow VTs to become the preferred architecture when compute resources are limited, and attempts to find a methodology that is recommended in these circumstances. It will, however, restrict itself to art classification tasks. Several experiments will be performed, of which results will be analysed both quantitatively as well as qualitatively.

%TODO: talk about datasets small enough to be hand labelled by one person
%The current study limits itself to the task of identifying which materials were used in artworks. For this, it will examine to what extent TL techniques (both fine-tuning and OTS) allow VTs to become the preferred architecture when compute resources are limited, and attempt to find a methodology that is preferred in circumstances as described. It will do this by means of several experiments, and both quantitative and qualitative analysis of the results.

More formally, the following research question is proposed: \begin{quote}\textit{To what extent (and in what manners) does utilizing TL allow VTs to perform on par with CNNs on common art classification problems, when compute resources are limited to a single GPU?}\end{quote}

### 2022-06-30 writing/tex/draft_methods.tex
Continued working on writing y paper

% Experiments build upon Sabatelli et al.
Much of the experiments and analysis discussed below, builds upon the work done in \citeauthor{sabatelli2018deep} (\citeyear{sabatelli2018deep}). This paper investigated TL properties of CNNs in the art classification domain. The main focus was on a comparison between CNN architectures pretrained on ImageNet1K when they were either used as OTS feature extractors, or fine-tuned to the new task. It was suggested that FT resulted in substantially better performance than either training from scratch or OTS TL.

@@

Similar to \citeauthor{sabatelli2018deep}, used models are pre-trained on ImageNet1K. Moreover, the \textit{Rijksmuseum Challenge} dataset \citep{mensink14icmr} is used to extract the target tasks from. This dataset consists of a large collection of digitized artworks, together with xml-formatted metadata. The classification tasks distilled from it are: (1) \textit{Type classification}, containing labels as `painting', `sculpture', `drawing', etc.; (2) \textit{Material classification}, with labels as `paper', `porcelain', `silver', etc.; and finally, (3) `Artist classification', where the model has to predict who the creator is.

@@

For robustness, five subsets are randomly generated per classification task. These are then split up into 80\%, 10\%, 10\% training, validating and testing sets, respectively.

Other than the tasks described above, table \ref{methods:datasets} shows an additional experiment called `Scaling'. The goal of this experiment is to see how the findings of this paper hold up as dataset sizes are shrunk further. It does this by taking the top 15 classes of type classification, and scaling it 4 times by a factor $\sqrt[4]{\frac{1}{10}} \approx 0.56$. The 5 datasets produced in this manner are then respectively 100\%, 56\%, 32\%, 18\% and 10\% the size of the dataset shown in table \ref{methods:datasets}. Steps are taken to ensure the distribution remains the same. Finally, 5 datasets are generated per scaling factor, leading to a total of 25 datasets. The same 80\%, 10\%, 10\% split is used for all.

@@

\subsection{Models} \label{methods:models}
Eight different neural network architectures are used throughout this paper, of which four are CNN-based and four ViT-based. When multiple versions are available (i.e. `base', `small', `tiny'), `base'/`medium' ones are chosen.

To start off with CNNs, ResNet50 \citep{he2016deep} and VGG19 \citep{simonyan2014very} are chosen, as these were the best performing models in \citeauthor{sabatelli2018deep} for respectively FT and OTS TL. In addition, \citeauthor{matsoukas2021time} (\citeyear{matsoukas2021time}) used ResNet50 too, while \citeauthor{zhou2021convnets} (\citeyear{zhou2021convnets}) used ResNet101 and ResNet152. To also add some more recent CNNs, ConvNext \citep{liu2022convnet} -- a purely CNN-based model inspired by ViTs' recent success -- and EfficientNetV2 \citep{tan2021efficientnetv2} are included as well.

For ViTs, versions with a 16x16 patch size are chosen. The first model is the original ViT \citep{dosovitskiy2020image}. Swin \citep{liu2021swin}, which was mentioned in section \ref{related_work}, is also selected, as it showed promising performance in \citeauthor{zhou2021convnets}. Section \ref{related_work} mentioned DeiT \citep{touvron2021training} too, as the distillation based architecture. The small version of this architecture was used in \citeauthor{matsoukas2021time} (\citeyear{matsoukas2021time}), since it is similar in size to ResNet50, to which it was compared. Here, however, a base-sized version is chosen. Specifically, the version that does not include output for distillation learning.
% TODO TODO TODO TODO TODO TODO TODO TODO YOU FORGOT TO MENTION BEIT!!!
% TODO if you do end up using distilation with ConvNext as a teacher: don't forget to change this section!!! TODO 2: im not completely happy with the part about DeiT. Doesn't flow super nicely

All ViT-based models have a bit over 86 million parameters. ConvNext is of similar size, with 88.6 million parameters. ResNet50 and EfficientNetV2 are smaller, with 25.6 and 13.6 million, respectively, while VGG19 is the largest model, with 143.7 million parameters. It should be mentioned that for both OTS TL and FT, preliminary experiments were performed with `small' and `tiny' models. The results of these suggested that similar conclusions could have been drawn if these were used instead.

% General information (4 per type, base/tiny etc)
% Paragraph/section about CNN models
%  - ResNet50 and VGG19 best fine-tuned and ots networks in Sabatelli
%  - ResNet50 I think also used in medical imaging paper; ResNet101 ots vit paper
%  * Convnext and EfficientNetV2 to have some more modern ones in the mix (to make comparison more fair)
% Paragraph/section about ViT models -- explain the idea behind each in 1 sentence maybe (also for CNNs)

@@

For all models, the linear classification layer is replaced as to fit the new task. Cross entropy loss uses the raw outputs of this layer (conform PyTorch documentation), together with one-hot encoded prediction targets. An early stop occurs after 10 epochs without improved loss on the validation set, and the model with the lowest loss is benchmarked on the testing set.

For OTS TL, the Adam optimizer is used with standard PyTorch parameters (\texttt{lr}=1e-3, $\beta_1=0.9$, $\beta_2=0.999$), and a batch size of 256. For FT, more is needed to achieve good convergence, especially for ViT-based architectures. Used hyperparameters, etc. are partially inspired by \citeauthor{matsoukas2021time} (\citeyear{matsoukas2021time}) and \citeauthor{zhou2021convnets} (\citeyear{zhou2021convnets}). The Adam learning rate now starts off at 1e-4, and is reduced by a factor 10 after 3 epochs without improvement (i.e. patience=2); the batch size is reduced to 32; label smoothing of 0.1 is added to the cross entropy loss, and a dropout layer with p=0.2 is inserted before the classification layer. Finally, input images are augmented with random horizontal flips and rotations in a $\pm 10 ^\circ$ range.

@@

The PyTorch machine learning framework is used for all experiments, and many pre-trained models are taken from its Torchvision library. Exceptions are EfficientNetV2, Swin, DeiT and Beit, which are taken from the Timm library\footnote{\url{https://timm.fast.ai/}.}.

>>>>

% Experiments build upon Sabatelli et al.
Much of the experiments and analysis discussed below, builds upon the work done in \citeauthor{sabatelli2018deep} (\citeyear{sabatelli2018deep}). This paper investigated TL properties of CNNs in the art classification domain. The main focus was on a comparison between CNN architectures pre-trained on ImageNet1K when they were either used as OTS feature extractors, or fine-tuned to the new task. It was suggested that FT resulted in substantially better performance than either training from scratch or OTS TL.

@@

\begin{figure*}[tb]
\includegraphics[width=\textwidth]{img/examples.png}
\caption{Example images from the \textit{Rijksmuseum Challenge}.}
\label{methods:eg}
\end{figure*}

Similar to \citeauthor{sabatelli2018deep}, used models are pre-trained on ImageNet1K. Moreover, the \textit{Rijksmuseum Challenge} dataset \citep{mensink14icmr} is used to extract the target tasks from. This dataset consists of a large collection of digitized artworks, together with xml-formatted metadata. The classification tasks distilled from it are: (1) \textit{Type classification}, containing labels as `painting', `sculpture', `drawing', etc.; (2) \textit{Material classification}, with labels as `paper', `porcelain', `silver', etc.; and finally, (3) \textit{Artist classification}, where the model has to predict who the creator is. Figure \ref{methods:eg} shows labelled examples for all three tasks.

@@

For robustness, five subsets are randomly generated per classification task. These are then split up into 80\%, 10\%, 10\% training, validating and testing sets, such that all experiments can be run five times. The average sample overlap between two of the five datasets is shown in table \ref{methods:datasets} as well.

Other than the tasks described above, table \ref{methods:datasets} shows an additional experiment called \textit{Scaling}. The goal of this experiment is to see how the findings of this paper hold up as dataset sizes are gradually shrunk. It does this by taking the top 15 classes of type classification, and scaling it 4 times by a factor $\sqrt[4]{\frac{1}{10}} \approx 0.56$. The 5 datasets produced in this manner are then respectively 100\%, 56\%, 32\%, 18\% and 10\% the size of the dataset shown in table \ref{methods:datasets}. Steps are taken to ensure the distribution remains the same. Finally, also here 5 datasets are generated per scaling factor, leading to a total of 25 datasets. The same 80\%, 10\%, 10\% split is used for all.

@@

\subsection{Models} \label{methods:models}
Eight different neural network architectures are used throughout this paper, of which four are CNN-based and four VT-based. When multiple versions are available (i.e. `base', `small', `tiny'), `base'/`medium' ones are chosen.

To start off with CNNs, \texttt{ResNet50} \citep{he2016deep} and \texttt{VGG19} \citep{simonyan2014very} are chosen, as these were the best performing models in \citeauthor{sabatelli2018deep} for respectively FT and OTS TL. In addition, \citeauthor{matsoukas2021time} (\citeyear{matsoukas2021time}) used \texttt{ResNet50} too, while \citeauthor{zhou2021convnets} (\citeyear{zhou2021convnets}) used \texttt{ResNet101} and \texttt{ResNet152}. To also add some more recent CNNs, \texttt{ConvNext} \citep{liu2022convnet} -- a purely CNN-based model inspired by VTs' recent success -- and \texttt{EfficientNetV2} \citep{tan2021efficientnetv2} are included as well.

For VTs, versions with a 16x16 patch size are chosen. The first model is the original VT \citep{dosovitskiy2020image}, which will be referred to as \texttt{ViT}. Next is \texttt{Swin} \citep{liu2021swin}, since it showed promising performance in \citeauthor{zhou2021convnets}. The final two are \texttt{DeiT} and \texttt{BeiT}. The small version of \texttt{DeiT} was used in \citeauthor{matsoukas2021time} (\citeyear{matsoukas2021time}), because it is similar in size to \texttt{ResNet50}, to which it was compared. In this work, however, also a base-sized version is chosen. Specifically, the version that does not include output for distillation learning.

% Section \ref{related_work} mentioned DeiT \citep{touvron2021training} too, as the distillation based architecture. The small version of this architecture was used in \citeauthor{matsoukas2021time} (\citeyear{matsoukas2021time}), since it is similar in size to ResNet50, to which it was compared. Here, however, a base-sized version is chosen. Specifically, the version that does not include output for distillation learning.
% TODO TODO TODO TODO TODO TODO TODO TODO YOU FORGOT TO MENTION BEIT!!!
% TODO 2: im not completely happy with the part about DeiT. Doesn't flow super nicely

All VT-based models have a bit over 86 million parameters. \texttt{ConvNext} is of similar size, with 88.6 million parameters. \texttt{ResNet50} and \texttt{EfficientNetV2} are smaller, with 25.6 and 13.6 million, respectively, while \texttt{VGG19} is the largest model, with 143.7 million parameters. It should be mentioned that for both OTS TL and FT, preliminary experiments were performed with `small' and `tiny' models. The results of these suggested that similar conclusions could have been drawn if these were used instead.

% General information (4 per type, base/tiny etc)
% Paragraph/section about CNN models
%  - ResNet50 and VGG19 best fine-tuned and ots networks in Sabatelli
%  - ResNet50 I think also used in medical imaging paper; ResNet101 ots vt paper
%  * Convnext and EfficientNetV2 to have some more modern ones in the mix (to make comparison more fair)
% Paragraph/section about VT models -- explain the idea behind each in 1 sentence maybe (also for CNNs)

@@

For all models, the linear classification layer is replaced as to fit the new task. Cross entropy loss uses the raw outputs of this layer (conform \texttt{PyTorch} documentation), together with one-hot encoded prediction targets. An early stop occurs after 10 epochs without improved loss on the validation set, and the model with the lowest loss is benchmarked on the testing set.

For OTS TL, the Adam optimizer is used with standard \texttt{PyTorch} parameters (\texttt{lr}=1e-3, $\beta_1=0.9$, $\beta_2=0.999$), and a batch size of 256. For FT, more is needed to achieve good convergence, especially for VT-based architectures. Used hyperparameters, etc. are partially inspired by \citeauthor{matsoukas2021time} (\citeyear{matsoukas2021time}) and \citeauthor{zhou2021convnets} (\citeyear{zhou2021convnets}). The Adam learning rate now starts off at 1e-4, and is reduced by a factor 10 after 3 epochs without improvement (i.e. patience=2); the batch size is reduced to 32; label smoothing of 0.1 is added to the cross entropy loss, and a dropout layer with p=0.2 is inserted before the classification layer. Finally, input images are augmented with random horizontal flips and rotations in a $\pm 10 ^\circ$ range.

@@

The \texttt{PyTorch} machine learning framework is used for all experiments, and many pre-trained models are taken from its \texttt{Torchvision} library. Exceptions are \texttt{EfficientNetV2}, \texttt{Swin}, \texttt{DeiT} and \texttt{BeiT}, which are taken from the \texttt{Timm} library\footnote{\url{https://timm.fast.ai/}.}.
